# Data Collection

## ABMI use of remote cameras

The ABMI has been using remote cameras to monitor mammal species in the province since 2014. 

As of `r Sys.Date()`, the ABMI has deployed X number of cameras across the province. 

## Camera deployment protocols

The ABMI's remote camera deployment protocols can be accessed here: [Terrestrial ABMI Autonomous Recording Unit (ARU) and Remote Camera Trap Protocols](https://abmi.ca/home/publications/551-600/599).

## List of ABMI camera projects

The following camera projects have been deployed by the ABMI:

Organize by year? Use gt package?

Note - would be nice to have a link to the WildTrax project for those projects that are published.

```{r, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}

library(wildRtrax)
library(keyring)
library(tidyverse)
library(kableExtra)

Sys.setenv(WT_USERNAME = key_get("WT_USERNAME", keyring = "wildtrax"),
           WT_PASSWORD = key_get("WT_PASSWORD", keyring = "wildtrax"))

wt_auth()

abmi_projects <- wt_get_download_summary(sensor_id = "CAM") |>
  filter(organization == "ABMI",
         !str_detect(project, "ACME|Pilot|Elk Island|Hare Coat|Sahtu|Ticks|Compression|wildRtrax|first|Lemming|Zoo")) |>
  mutate(Project = str_remove(project, "^ABMI ")) |>
  select(Project, `No. Cameras` = tasks, Status = status)
  # Would it be better to arrange by year?
  # arrange(desc(Cameras))

kable(abmi_projects)

```


### Image Tagging

Image tagging is done on the [WildTrax](https://wildtrax.ca/about/) platform. Raw data is also publicly available from the ABMI via this platform.

The image tagging protocols used by the ABMI can be accessed here: 

There is a companion R package, [*wildRtrax*](https://abbiodiversity.github.io/wildRtrax/), which is also publicly available and contains many useful functions for downloading and analyzing camera data hosted on the platform. Throughout this document, references will be made to functions within this package.

Note: Would be nice to insert the wildRtrax package logo. 

## Species detections (2014-2021)

The map below displays mammal species detections at ABMI Ecosystem Health (core grid) sites sampled with cameras between 2014 and 2023. Each site is sampled with 4 individual cameras placed 600-m apart in a square shape.

(Note: Need to add EH 2023)

```{r, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}

library(sf)
library(crosstalk) # Special branch of GH crosstalk repo required (filter-select-default)
library(leaflet)
library(leaflet.extras)
 
g_drive <- "G:/Shared drives/ABMI Camera Mammals/"

# Alberta provincial polygon
sf_ab <- st_read(paste0(g_drive, "data/spatial/ab_prov.shp"),
                 quiet = TRUE, stringsAsFactors = FALSE)

# Site locations
sites <- read_csv(paste0(g_drive, "data/lookup/climate/site-climate-summary_v2020.csv")) |>
  select(site = SITE_ID,
         latitude = PUBLIC_LATTITUDE,
         longitude = PUBLIC_LONGITUDE)

# Occurrence data
occ <- read_csv(paste0(g_drive, "Results/Occurrence/ABMI/occurrence_eh_site_2023-01-23.csv")) |>
  filter(!str_detect(site, "^W")) |>
  mutate(site = as.numeric(site))

# Species to display
sp <- occ |>
  group_by(common_name) |>
  summarise(n = sum(present)) |>
  filter(n >= 35,
         !str_detect(common_name, "Ermine|Squirrel"),
         !common_name == "Deer") |>
  select(common_name) |>
  pull()

occ_subset <- occ |>
  filter(common_name %in% sp) |>
  mutate(year = str_sub(project, start = -4, end = -1)) |>
  full_join(sites, by = "site") |>
  mutate(present = ifelse(is.na(present), "Not Surveyed", present)) |>
  mutate(present = factor(present, levels = c("Not Surveyed",
                                                 "TRUE",
                                                 "FALSE"))) |>
  select(site, latitude, longitude, year, common_name, present) |>
  arrange(site, year)

# Create SharedData object for crosstalk interactivity
shared_ss <- SharedData$new(occ_subset)

# Palette
pal <- colorFactor(palette = c("grey80",
                               "gold",
                               "black"),
                   domain = occ_subset$present)

# Create leaflet map
map <- sf_ab |>
  st_transform("+init=epsg:4326") %>%
  leaflet() %>%
  addTiles() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addFullscreenControl() %>%
  addResetMapButton() %>%
  addScaleBar(position = "bottomright", 
              options = scaleBarOptions(imperial = FALSE)) |>
  
  addPolylines(color = "#0a0909", weight = 2.5, smoothFactor = 0.2, opacity = 3, fill = FALSE) |>
  
  addCircleMarkers(data = shared_ss,
                   color = ~pal(present), stroke = FALSE, fillOpacity = 1,
                   radius = 6) |>
  
  addLegend(data = shared_ss, pal = pal, values = ~present,
            position = "topright", opacity = 1, title = "Occurrence at Site:")

```

```{r, out.width="100%", fig.height=10, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}

# Create interactive filters
bscols(widths = c(3, 3),
       filter_checkbox("year", "Sampling Year:", shared_ss, ~year, columns = 2),
       filter_select("common_name", "Species:", shared_ss, ~common_name, multiple = FALSE, selected = "Badger"))

# Display map
map

```


## Deployment duration lengths

Here we can include all those figures. Yay. Do we really want that?





